openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-10-01 paper/policar2020-opentsne-BMC.tex
More edits to main text

\documentclass[twocolumn]{article}

@@

\usepackage{amsmath,amssymb}
\usepackage{booktabs}

\newcommand{\opentsne}{\textsf{openTSNE}}

@@

@@

\author{
  Pavlin G. Poli\v{c}ar\textsuperscript{1}\\
  \texttt{pavlin.policar@fri.uni-lj.si}\\
  \and
  Martin Stra\v{zar}\textsuperscript{2}\\
  \and
  Bla\v{z} Zupan\textsuperscript{1,3}\\
}

\date{
  \textsuperscript{1}Faculty of Computer and Information Science, University of Ljubljana,\\
  Ve\v{c}na pot 113, SI-1000 Ljubljana, Slovenia\\ \vspace{2mm}
  \textsuperscript{2}Broad Institute of Harvard and MIT,\\
  D\"{u}sternbrooker Weg 20, MA-02142 Cambridge, U.S.A.\\ \vspace{2mm}
  \textsuperscript{3}Department of Molecular and Human Genetics, Baylor College of Medicine,\\
  1 Baylor Plaza, TX-77030 Houston, U.S.A.
}

\maketitle

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology can reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its
extensions. The library is orders of magnitude faster than existing
popular implementations, including those from scikit-learn. Unique to
\opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

\providecommand{\keywords}[1]{\vspace{2mm}\textbf{Keywords ---} #1}
\begin{keywords}
t-SNE, embedding, visualization, dimensionality reduction, Python library
\end{keywords}

\section*{Introduction}

The abundance of high-dimensional data sets in molecular biology calls for
dimensionality reduction techniques, particularly for methods that produce
informative data visualizations. Popular approaches include principal component
analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor
embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE
lately received much attention as it can address high volumes of data and reveal
a meaningful clustering structure. Increased resolution and throughput of modern
molecular assays has lead to the frequent use of t-SNE in diverse fields
including, but not limited to, single-cell
transcriptomics (scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE embeds high-dimensional expression profiles into a
two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
two such embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
  \caption{\label{fig:macosko}
  We use \opentsne\ to generate three t-SNE embeddings and demonstrate recent
  theoretical advances. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
  single-cell gene-expression profiles of mouse retinal cells from Macosko
  \textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
  additionally contains 27,499 expression profiles from mouse
  retinal cells from Shekhar \textit{et al.}~\cite{shekhar2016comprehensive}.
  \textbf{(a)} We construct a t-SNE
  embedding following the parameter choices from the original publication
  by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
  shows no preservation of the global organization of clusters,
  resulting from random initialization and an affinity model focused on
  preserving local neighborhoods. \textbf{(b)} A modern t-SNE
  embedding, utilizing the latest theoretical advances and practical
  recommendations constructed using a multi-scale
  affinity model, preserving both short-range and long-range interactions
  between data points and initialized so that the global layout is
  as meaningful as possible. Unlike in \textbf{(a)}, the green and blue clusters
  representing different sub-types of amacrine and bipolar cells are now
  localized to the same regions of the space, indicating a higher level
  of similarity than to other cell types. The embedding in \textbf{(c)} shows how existing
  t-SNE reference atlases can be used to place new samples into existing
  embeddings. The positions of new data points correspond to cell types
  from the reference atlas.
}
\end{figure*}

@@

\opentsne\ is compatible with the Python data science ecosystem and
libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
thus providing a familiar and intuitive application program interface to the new
users. Its modular design encourages extensibility and experimentation with
various settings and changes in the analysis pipeline.  We have released
\opentsne\ in an open-source to support adaptations by the wider community. At
the time of the writing, the library averages thousands of weekly
downloads and has received almost 700 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

@@

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use \opentsne\ to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
	each providing a different perspective into the topology of the data.
	The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent the cluster absorption. We
constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction
between these two depictions reveals that many clusters are composed of
subgroups representing different cell subtypes.

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
  Increasing the exaggeration parameter leads to compact clusters, highlighting
	the data's global organization and emphasizing the continuous nature of
	cell state transitions. The data set from Cao \textit{et al.}~\cite{cao2019single}
	contains expression profiles from 2,058,652 single cells. The data were
	collected from mice embryos at different developmental stages at daily
	intervals after 9.5 to 13.5 days. \textbf{(c)} reveals that the data is
	comprised of two main components -- the neural tube and mesenchymal
	cells -- as well as several other smaller clusters. The
	colors indicate developmental progression with red indicating
	least-developed cells and blue indicating most developed cells. The
	overall developmental trajectory is most apparent with higher
	exaggeration levels, showing red cells slowly transitioning into blue
	cells. Progressively easing the exaggeration factor uncovers finer
	clusters within the larger groups, as shown in \textbf{(b)} with
	exaggeration of two and subsequently in \textbf{(a)}, where we show the
	standard t-SNE with no exaggeration. 32,011 putative doublets are
	excluded from the visualizations.
}
\end{figure*}

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is changed
continuously. To this end, other embedding techniques such as UMAP and
ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity
between cell states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying}
have shown that embeddings produced by t-SNE with exaggeration values of 4 and
$\sim30$ construct embeddings which are markedly similar to UMAP and
ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the
developmental trajectory between different time points is difficult to observe
due to many sprawled out clusters. On the other hand, it is easier to trace the
development when we increase the exaggeration factor from $1$ to $4$
(Fig.~\ref{fig:cao}.b-c).

@@

Versatility, the ability to use and combine different optimization approaches to
construct the embedding spaces, is another of \opentsne's core design
principles. Kobak \& Berens recently introduced several recommendations and
tricks to obtain better and more meaningful t-SNE
visualizations~\cite{kobak2019art}. These include multi-scale similarity
kernels, perplexity annealing, and increasing exaggeration when working with
more massive data sets. \opentsne\ provides a flexible program interface to
incorporate these improvements in just a few lines of code. Furthermore, \opentsne\
supports custom affinity models, enabling users to construct t-SNE embeddings on
non-tabular relational data: the only requirement imposed by the affinity-model
is the availability of similarities between data points.
Finally, \opentsne's comprehensive callback system can be utilized
to monitor and adapt different stages of the optimization phase and has been
used to construct visually appealing animations of the t-SNE optimization
process.

@@

One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\cite{becht2019dimensionality}. The culprit for slow response time stems
from a particular optimization procedure and its specific implementation in
popular Python libraries. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
e.g.} cells). The most widely-used implementation of t-SNE came from
\textsf{scikit-learn}, which exhibits long runtimes when compared to its C++
counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE}
implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in
\opentsne\, enabling the embedding of large data sets in a matter of minutes.

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}
  We benchmark \opentsne\ (v0.4.3) against three popular open-source implementations
	from \textsf{scikit-learn}~\cite{pedregosa2011scikit} (v0.23.1),
	\textsf{MulticoreTSNE}~\cite{Ulyanov2016} (v0.1), and
	\textsf{FIt-SNE}~\cite{linderman2019fast} (v1.1.0). Experiments were run on a
	consumer-grade Intel Core i7-7700HQ processor found in laptop computers,
	and on a server-grade Intel Xeon E5-2650. To generate benchmark data
	sets of different sizes, we subsampled data from the 10X Genomics 1.3
	million mouse brain data set five times, resulting in five different
	data sets for each size. In total, we run each implementation on 30
	different data sets. Notice that \opentsne\ scales similarly to \textsf{FIt-SNE},
	as they both use the same interpolation-based approximation scheme,
	while \textsf{scikit-learn} and \textsf{MulticoreTSNE} utilize the Barnes-Hut
	approximation.
}
\end{figure*}

Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \textsf{FIt-SNE}
and \opentsne\ scale better than their Barnes-Hut counterparts --
\textsf{scikit-learn} and \textsf{MulticoreTSNE}. While \opentsne's
implementation uses Python that incurs some runtime overhead compared to its C++
counterpart \textsf{FIt-SNE}, the two libraries' speeds are surprisingly
comparable. Modern computer processors contain multiple cores, allowing us to
use multi-threading, which further reduces the gap between \opentsne\ and
\textsf{FIt-SNE}. On the server-grade processor, \opentsne\ is slightly faster
than its pure C++ counterpart when utilizing multiple cores. \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which is aggressively optimized on Intel(R)
Xeon(R) processors.

@@

We have designed \opentsne\ to make it accessible to a broader audience. We
provide precompiled binaries for all major Python versions on all major
platforms, making the installation process as seamless as possible. One can
install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} from the
\textsf{conda-forge} channel. \opentsne's interface is inspired by \textsf{scikit-learn},
which is well established in the Python data science ecosystem. \opentsne\
implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
approximation schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance overhead,
the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
\opentsne\ is extensible. Its modular design enables researchers to quickly
experiment with different parameter settings and easily incorporate their own components into
the software. We provide a full feature list and comparison to other popular
t-SNE implementations in Table~\ref{tab:features}.

\begin{table}
\caption{\label{tab:features}
  We compare features of \opentsne\ to three popular open-source implementations
	from \textsf{scikit-learn}~(v0.23.1), \textsf{MulticoreTSNE} (v0.1), and
	\textsf{FIt-SNE} (v1.1.0).  The first section of the comparison
	addresses packaging and distribution. A properly packaged library is
	easily accessible to users, and developers should easily include it in
	dependency lists of other software packages. The second section of the
	comparison lists the two existing t-SNE approximation schemes. The
	FIt-SNE approximation scheme is required for t-SNE to scale up to
	millions of data points. The final section provides a list of extensions
	and improvements to the standard t-SNE algorithm, many of which can
	produce markedly better visualizations.
}

@@

\begin{tabular}{r c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
\toprule
\textsf{PyPI} package & \checkmark & \checkmark & & \checkmark \\
\textsf{conda} package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
\hline
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & \checkmark \\
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Better initialization & & & \checkmark & \checkmark \\
Automatic learning rate & & & \checkmark & \checkmark \\
Embedding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

@@

\fontsize{8}{10}\selectfont
\section*{Methods}

@@

@@

\end{document}

>>>>

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

\documentclass[twocolumn]{bmcart}
%\documentclass{bmcart}

@@

\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\startlocaldefs
\newcommand{\opentsne}{\textsf{openTSNE}}
\endlocaldefs

@@

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Software}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={pavlin.policar@fri.uni-lj.si}   % email address
]{\inits{PGP}\fnm{Pavlin G.} \snm{Poli\v{c}ar}}
\author[
   addressref={aff1,aff2}
]{\inits{MS}\fnm{Martin} \snm{Stra\v{z}ar}}
\author[
   addressref={aff1,aff3}
]{\inits{BZ}\fnm{Bla\v{z}} \snm{Zupan}}

\address[id=aff1]{%
  \orgname{Faculty of Computer and Information Science, University of Ljubljana},
  %\street{Ve\v{c}na Pot 113},
  \postcode{SI 1000}
  \city{Ljubljana},
  \cny{Slovenia}
}
\address[id=aff2]{%
  \orgname{Broad Institute of Harvard and MIT},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{MA 02142}
  \city{Cambridge},
  \cny{U.S.A.}
}
\address[id=aff3]{%
  \orgname{Department of Molecular and Human Genetics, Baylor College of Medicine},
  %\street{D\"{u}sternbrooker Weg 20},
  \postcode{TX 77030}
  \city{Houston},
  \cny{U.S.A.}
}

\end{fmbox}% comment this for two column layout

\begin{abstractbox}
\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology can reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its
extensions. The library is orders of magnitude faster than existing
popular implementations, including those from scikit-learn. Unique to
\opentsne\ is also mapping new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

\begin{keyword}
\kwd{t-SNE}
\kwd{embedding}
\kwd{visualization}
\kwd{dimensionality reduction}
\kwd{Python library}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

\section*{Background}

The abundance of high-dimensional data sets in molecular biology calls for
dimensionality reduction techniques, particularly for methods that produce
informative data visualizations. Popular approaches include principal component
analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor
embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE
lately received much attention as it can address high volumes of data and reveal
a meaningful clustering structure. Increased resolution and throughput of modern
molecular assays has lead to the frequent use of t-SNE in diverse fields
including, but not limited to, single-cell
transcriptomics(scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE embeds high-dimensional expression profiles into a
two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
two such embeddings.

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{macosko2015}
\caption{\label{fig:macosko}
Three t-SNE embeddings demonstrate theoretical advances and corresponding implementations in
\opentsne. The data in \textbf{(a)} and \textbf{(b)} represent 44,808
single-cell gene-expression profiles of mouse retinal cells from Macosko
\textit{et al.}~\cite{macosko2015highly}. The data in \textbf{(c)}
contains 27,499 expression profiles from mouse
retinal cells from Shekhar \textit{et
al.}~\cite{shekhar2016comprehensive}. \textbf{(a)} We construct a t-SNE
embedding following the parameter choices from the original publication
by Maaten \& Hinton~\cite{maaten2008visualizing}. The visualization
shows no preservation of the global organization of clusters,
resulting from random initialization and an affinity model focused on
preserving local neighborhoods. \textbf{(b)} A modern t-SNE
embedding, utilizing the latest theoretical advances and practical
recommendations constructed using a multi-scale
affinity model, preserving both short-range and long-range interactions
between data points and initialized so that the global layout is
meaningful. Unlike in \textbf{(a)}, the green and blue clusters
representing different sub-types of amacrine and bipolar cells are now
localized to the same regions of the space, indicating a higher level
of similarity than to other cell types. Embedding in \textbf{(c)} shows how existing
t-SNE reference atlases can be used to place new samples into existing
embeddings. The positions of new data points correspond to cell types
from the reference atlas.
}
\end{figure*}

@@

The \opentsne\ is compatible with the Python data science ecosystem and
libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
thus providing a familiar and intuitive application program interface to the new
users. Its modular design encourages extensibility and experimentation with
various settings and changes in the analysis pipeline.  We have released
\opentsne\ in an open-source to support adaptations by the wider community. At
the time of the writing, the library averages several thousands of weekly
downloads and has received almost 700 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

@@

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use \opentsne\ to create four different
    visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
    each providing a different perspective into the topology of the data.
    The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. The standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a good trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
	smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. Embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters by the larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent the cluster absorption. We
constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction
between these two depictions reveals that many clusters are composed of
subgroups representing different cell subtypes.

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{cao2019}
  \caption{\label{fig:cao}
Increasing the exaggeration parameter leads to compact clusters, highlighting the data's global organization and emphasizing the continuity between cell states. The data set from Cao \textit{et al.}~\cite{cao2019single} contains expression profiles from 2,058,652 single cells. The data were collected from mice embryos at different development stages at daily intervals after 9.5 to 13.5 days. \textbf{(a)} reveals that the data is comprised of two main components -- the neural tube and mesenchymal cells -- as well as several other smaller clusters. Additionally, the colors indicate developmental progression with red indicating least-developed cells and blue indicating most developed cells. The overall developmental trajectory is most apparent with higher exaggeration levels, showing red cells slowly transitioning into blue cells. Progressively easing the exaggeration factor uncovers finer clusters within the larger groups, as shown in
\textbf{(b)} with exaggeration of two and subsequently in \textbf{(c)}, where we show the standard t-SNE with no exaggeration. 32,011 putative doublets are excluded from the visualizations.
}
\end{figure*}

Exaggeration can highlight transitions between cell states in developmental studies. Standard t-SNE often produces embeddings with clearly defined, discrete clusters. We can adjust the level of granularity and resolution of the clusters with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are often undesired in developmental studies where cells' state is changed continuously. To this end, other embedding techniques such as UMAP and ForceAtlas2~\cite{jacomy2014forceatlas2} tend to capture the continuity between cell states better. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying} have shown that embeddings produced by t-SNE with exaggeration values of 4 and $\sim30$ construct embeddings which are markedly similar to UMAP and ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the developmental trajectory between different time points is difficult to observe due to many sprawled out clusters. On the other hand, it is easier to trace the development when we increase the exaggeration factor from $1$ to $4$ (Fig.~\ref{fig:cao}.b-c).

@@

Versatility, the ability to use and combine different optimization approaches to
construct the embedding spaces, is another of \opentsne's core design
principles. Kobak \& Berens recently introduced several recommendations and
tricks to obtain better and more meaningful t-SNE
visualizations~\cite{kobak2019art}. These include multi-scale similarity
kernels, perplexity annealing, and increasing exaggeration when working with
more massive data sets. \opentsne\ provides a flexible program interface to
incorporate these improvements in just a few code lines. Furthermore, \opentsne\
supports custom affinity models, enabling users to construct t-SNE embeddings on
non-tabular relational data: the only requirement imposed by the affinity-model
is the availability of similarity between data points, that is, the distance
matrix.  Finally, the \opentsne's comprehensive callback system can be utilized
to monitor and adapt different stages of the optimization phase and has been
used to construct visually appealing animations of the t-SNE optimization
process.

@@

One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\cite{becht2019dimensionality}. The culprit for slow response time stems
from a particular optimization procedure and its specific implementation in the
popular Python library. Namely, until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
e.g.} cells). The most widely-used implementation of t-SNE came from
\textsf{scikit-learn}, which exhibits long runtimes when compared to its C++
counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE}
implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in
\opentsne\, enabling the embedding of large data sets in a matter of minutes.

@@

\begin{figure*}[htbp]
  \includegraphics[width=\textwidth]{benchmarks}
  \caption{\label{fig:benchmarks}
We benchmark openTSNE (v0.4.3) against three popular open-source implementations
	from scikit-learn~\cite{pedregosa2011scikit} (v0.23.1),
	MulticoreTSNE~\cite{Ulyanov2016} (v0.1), and
	FIt-SNE~\cite{linderman2019fast} (v1.1.0). Experiments were run on a
	consumer-grade Intel Core i7-7700HQ processor found in laptop computers,
	and on a server-grade Intel Xeon E5-2650. To generate benchmark data
	sets of different sizes, we subsampled data from the 10X Genomics 1.3
	million mouse brain data set five times, resulting in five different
	data sets for each size. In total, we run each implementation on 30
	different data sets. Notice that openTSNE scales similarly to FIt-SNE,
	as they both use the same interpolation-based approximation scheme,
	while scikit-learn and MulticoreTSNE utilize the Barnes-Hut
	approximation.
}
\end{figure*}

Benchmark results (Fig.~\ref{fig:benchmarks}) confirm that both \textsf{FIt-SNE}
and \opentsne\ scale better than their Barnes-Hut counterparts --
\textsf{scikit-learn} and \textsf{MulticoreTSNE}. While \opentsne\'s
implementation uses Python that incurs some runtime overhead compared to its C++
counterpart \textsf{FIt-SNE}, the two libraries' speeds are surprisingly
comparable. Modern computer processors contain multiple cores, allowing us to
use multi-threading, which further reduces the gap between \opentsne\ and
\textsf{FIt-SNE}. On the server-grade processor, \opentsne\ is slightly faster
than its pure C++ counterpart when utilizing multiple cores. \opentsne\ uses
\textsf{numpy} for most linear algebra operations, which may make better use of
the Intel Math Kernel Library (MKL), which is aggressively optimized on Intel(R)
Xeon(R) processors.

@@

We have designed \opentsne\ to make it accessible to a broader audience. We
provide precompiled binaries for all major Python versions on all major
platforms, making the installation process as seamless as possible. One can
install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} on the
\textsf{conda-forge} channel. \opentsne\ interface follows that of \textsf{scikit-learn},
which is well established in the Python data science ecosystem. \opentsne\
implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
approximation schemes, enabling it for data sets containing millions of data
points. While the Python virtual machine introduces some performance overhead,
the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
\opentsne\ is extensible. Its modular design enables researchers to quickly
experiment with different settings and easily incorporate their components into
the software. We provide a full feature list and comparison to other popular
t-SNE implementations in Table~\ref{tab:features}.

\begin{table}
\caption{\label{tab:features}
We compare features of \opentsne to three popular open-source implementations
	from scikit-learn~(v0.23.1), MulticoreTSNE (v0.1), and FIt-SNE (v1.1.0).
	The first section of the comparison addresses packaging and
	distribution. A properly packaged library is easily accessible to users,
	and developers should easily include it in dependency lists of other
	software packages. The second section of the comparison lists the two
	existing t-SNE approximation schemes. The FIt-SNE approximation scheme
	is required for t-SNE to scale up to millions of data points. The final
	section provides a list of extensions and improvements to the standard
	t-SNE algorithm, many of which can produce markedly better
	visualizations.
}

@@

\begin{tabular}{r c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
\toprule
PyPI package & \checkmark & \checkmark & & \checkmark \\
conda package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
\hline
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & \checkmark \\
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Better initialization & & & \checkmark & \checkmark \\
Automatic learning rate & & & \checkmark & \checkmark \\
Embedding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

@@

\section*{Methods}

@@

\begin{backmatter}

@@

\end{backmatter}
\end{document}
